LOINC2HPO_EPIC AND SCC

### 2022-03-11 src/main/r/MSDW_HPO_Phenotype_Analysis.Rmd
update analysis

@@

  if (force_rerun | !file.exists(df_path)){
    df <- dbGetQuery(dsca, "
            select race, gender, count(*) as n
            from hai_az_prod.person_msdw_2020july pmj 
            group by race, gender 
            order by race, gender;
            ")

@@

patient_race_sex_distribution <- get_patient_race_sex_distribution()

@@

# recode "African American" as "Black or African American"
# TODO: update the race map
patient_race_sex_distribution <- patient_race_sex_distribution %>%
  mutate(race = recode(race, 'African American' = 'Black or African American'))
```

@@

ggsave(filename = glue('{CACHE_DIR}/plot_by_sex.png'), plot = plot_by_sex, width = 2.5, height = 2.5)
plot_by_sex

@@

  if (force_rerun | !file.exists(df_path)){
    df <- dbGetQuery(dsca, "
            with person_birth_year as (
            	select medical_record_number , date_of_birth , gender , date_part('Year', date_of_birth) as birth_year 
            	from hai_az_prod.person_msdw_2020july pmj
            )
            select birth_year, gender, count(*) as n 
            from person_birth_year 
            group by birth_year , gender 
            order by birth_year, gender;
            ")

@@

patient_birth_year_sex_distribution <- get_patient_birth_year_distribution()

@@

### Count patient distribution by the number of HPO terms per patient 
```{r}

>>>>

### Identify starting cohort
The starting cohort could be the entirety of the MSDW set, but here we focus on the subcohort where we have lab data for. 
```{sql connection=dsca}
create temp table cohort as 
select distinct medical_record_number 
from hai_az_prod.lab_scc_2020july lsj 
union 
select distinct mrn as medical_record_number 
from hai_az_test.lab_epic_2020july lej ;
```

@@

  if (force_rerun | !file.exists(df_path)){
    df <- dbGetQuery(dsca, "
            select race, gender, count(*) as n
            from hai_az_prod.person_msdw_2020july pmj 
            join cohort using (medical_record_number)
            group by race, gender 
            order by race, gender;
            ")

@@

patient_race_sex_distribution <- get_patient_race_sex_distribution(force_rerun = FALSE)

@@

# the following cleaning are no longer needed after updating the race mapping file
# recode "African American" as "Black or African American"
# patient_race_sex_distribution <- patient_race_sex_distribution %>%
#   mutate(race = recode(race, 'African American' = 'Black or African American'))
```

@@

ggsave(filename = glue('{CACHE_DIR}/plot_by_sex.png'), plot = plot_by_sex, width = 2.5, height = 2.5)
plot_by_sex
```

@@

  if (force_rerun | !file.exists(df_path)){
    df <- dbGetQuery(dsca, "
            with person_birth_year as (
            	select medical_record_number , date_of_birth , gender , date_part('Year', date_of_birth) as birth_year 
            	from hai_az_prod.person_msdw_2020july pmj
            )
            select birth_year, gender, count(*) as n 
            from person_birth_year 
            join cohort using (medical_record_number)
            group by birth_year , gender 
            order by birth_year, gender;
            ")

@@

patient_birth_year_sex_distribution <- get_patient_birth_year_distribution(FALSE)

@@

### Success rate of LOINC2HPO
```{r}
# among all the labs, what is the percentage that can be mapped to HPO
get_loinc2hpo_success_metric <- function(force_rerun=FALSE, cache_dir=CACHE_DIR){
  df_path = glue('{cache_dir}/loinc2hpo_success_metric.csv')

  if (force_rerun | !file.exists(df_path)){
    df <- dbGetQuery(dsca, "
            with scc_lab_count as (
            	select context_name , count(*) as n_test
            	from hai_az_prod.lab_scc_2020july lsj 
            	join hai_az_prod.fd_procedure fp using (procedure_key)
            	group by context_name
            ),
            epic_lab as (
            	select distinct mrn, order_date, lab_time , test_code, test_result_value  from prod_msdw.epic_lab el
            ),
            epic_lab_count as (
            	select 'EPIC' as context_name, count(*) as n_test from epic_lab
            ),
            lab_count as (
            	select * from scc_lab_count 
            	union all 
            	select * from epic_lab_count 
            ),
            loinc_success_count as (
            	select context_name , count(*) as n_loinc_success
            	from hai_az_prod.scc_and_epic_lab_after_loinc2hpo_2020july saelalhj 
            	where loinc is not null
            	group by context_name
            ),
            loinc2hpo_success_count as (
            	select context_name , count(*) as n_loinc2hpo_success
            	from hai_az_prod.scc_and_epic_lab_after_loinc2hpo_2020july saelalhj 
            	where hpotermid is not null
            	group by context_name 
            )
            select *, (n_loinc_success + 0.1)/n_test as loinc_success_rate, (n_loinc2hpo_success + 0.1)/n_test as loinc2hpo_success_rate
            from lab_count
            left join loinc_success_count using (context_name)
            left join loinc2hpo_success_count using (context_name);
            ")

loinc2hpo_success_metric <- get_loinc2hpo_success_metric(TRUE)

loinc2hpo_success_metric <- loinc2hpo_success_metric %>% mutate(across(function(x) is.numeric(x), function(x) replace_na(x, 0)))

loinc2hpo_success_metric %>% mutate(across(function(x) is.double(x), function(x) round(x, digits = 3))) %>%
  arrange(-n_test)
```
The overall success rate : `sum(loinc2hpo_success_metric$n_success) / sum(loinc2hpo_success_metric$n_test)`
```{r}
overall_success_rate <- sum(loinc2hpo_success_metric$n_success) / sum(loinc2hpo_success_metric$n_test)
print(glue('Total Lab Results: {sum(loinc2hpo_success_metric$n_test)}, 
           Total Lab Mapped to LOINC: {sum(loinc2hpo_success_metric$n_loinc_success)}
           Total Lab Transformed to HPO: {sum(loinc2hpo_success_metric$n_loinc2hpo_success)}'))

print(glue('Total LOINC Mapping Rate: {sum(loinc2hpo_success_metric$n_loinc_success)/sum(loinc2hpo_success_metric$n_test)},
           Total HPO transformation Rate: {sum(loinc2hpo_success_metric$n_loinc2hpo_success)/sum(loinc2hpo_success_metric$n_test)}'))
```

### Count patient distribution by the number of HPO terms per patient 
```{r}
# prepare a summary table: summary_hpo_counts_per_patient 
# rows: per patient
# columns: medical_record_number, hpo, unique hpo, abnormal_hpo, unique_abnormal_hpo, unique_abnormal_hpo_inference

# collect statistics for the starting cohort
# The mapping procedure assigned an average of 633 laboratory test-derived HPO terms per individual patient, many of which were from the same laboratory tests performed at different visits. The tests corresponded to a mean of 57.7 unique HPO terms, of which 20.8 were abnormalities and the remainder were normal phenotypes (Fig. 4e). The hierarchical structure of the HPO allows inferences to be propagated up to parent terms and their ancestors;14 using this method, we inferred an additional 51.2 HPO terms (total 73.5) based on 22.2 abnormalities for each patient
# how many phenotypes are mapped from lab tests? # unique phenotypes? then break down to normal and abnormal
# how many unique abnormal phenotypes for each patient? then after HPO inference

get_hpo_counts_per_patient <- function(force_rerun=FALSE, cache_dir=CACHE_DIR){
  df_path = glue('{cache_dir}/hpo_counts_per_patient.csv')

  if (force_rerun | !file.exists(df_path)){
    df <- dbGetQuery(dsca, "
            with count_no_inference as (
            	select medical_record_number , count(*) as n_hpo, count(distinct case when isnegated then 'NOT' else '' end || hpotermid) as n_distinct_hpo
            	from hai_az_prod.scc_and_epic_lab_after_loinc2hpo_2020july lab 
            	where hpotermid is not null
            	group by medical_record_number  
            ),
            count_abnormal_no_inference as (
            	select medical_record_number , count(distinct hpotermid) as n_distinct_abnormal_hpo
            	from hai_az_prod.scc_and_epic_lab_after_loinc2hpo_2020july lab 
            	where hpotermid is not null and not isnegated 
            	group by medical_record_number
            ),
            count_abnormal_with_inference as (
            	select medical_record_number , count(distinct ancestor) as n_distinct_abnormal_hpo_infer
            	from hai_az_prod.scc_and_epic_lab_after_loinc2hpo_2020july lab 
            	join hai_az_prod.hpo_is_a_pairs hiap on lab.hpotermid = hiap.current 
            	where hpotermid is not null and not isnegated 
            	group by medical_record_number
            )
            select c1.n_hpo, c1.n_distinct_hpo, c2.n_distinct_abnormal_hpo, c3.n_distinct_abnormal_hpo_infer 
            from cohort p 
            left join count_no_inference c1 using (medical_record_number) 
            left join count_abnormal_no_inference c2 using (medical_record_number)
            left join count_abnormal_with_inference c3 using (medical_record_number);
            ")

hpo_counts_per_patient <- get_hpo_counts_per_patient(TRUE)
hpo_counts_per_patient <- hpo_counts_per_patient %>% 
  replace_na(list(n_hpo = 0, n_distinct_hpo = 0, n_distinct_abnormal_hpo = 0, n_distinct_abnormal_hpo_infer = 0))
```

Report out the patient level phenotype counts
```{r}
# mean
sapply(hpo_counts_per_patient, mean)

# median 
sapply(hpo_counts_per_patient, median)
```

Plot the distribution of distinct phenotypes for each patient (including abnormal and abnormal)
```{r}
plot_by_phenotypes_per_patient <- ggplot(hpo_counts_per_patient) +
  geom_histogram(aes(x = n_distinct_hpo, y = ..count../1000), binwidth = 10, fill = 'orange', color = 'black', size = 0.25) +
  xlab('number of unique phenotypes for each patient') +
  ylab('patient counts (in thousands)') + 
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.title = element_text(size = 8), axis.text = element_text(size = 8), 
        legend.text = element_text(size = 8), legend.title = element_text(size = 10))
ggsave(filename = glue("{CACHE_DIR}/plotByPhenotypesPerPatient.png"), plot = plot_by_phenotypes_per_patient, width = 3, height = 2)
plot_by_phenotypes_per_patient
```
Plot distribution of abnormal phenotypes (distinct) after inference
```{r}
plot_by_distinct_abnormal_hpo_infer_per_patient <- ggplot(hpo_counts_per_patient) +
  geom_histogram(aes(x = n_distinct_abnormal_hpo_infer, y = ..count../1000), binwidth = 10, fill = 'orange', color = 'black', size = 0.25) +
  xlab('number of unique phenotypes for each patient') +
  ylab('patient counts (in thousands)') + 
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.title = element_text(size = 8), axis.text = element_text(size = 8), 
        legend.text = element_text(size = 8), legend.title = element_text(size = 10))
ggsave(filename = glue("{CACHE_DIR}/plotByDistinctAbnormalPhenotypeInferPerPatient.png"), plot = plot_by_distinct_abnormal_hpo_infer_per_patient, width = 3, height = 2)
plot_by_distinct_abnormal_hpo_infer_per_patient
```
